Project Protocol

### 2021-01-11 04-experiment.Rmd

---
title: "Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics"
affiliation:
  #author-columnar: true         ## one column per author
  institution-columnar: true  ## one column per institution (multiple autors eventually)
  # wide: true                  ## one column wide author/affiliation fields
  institution:
    - name: Queensland University of Technology
      department: Science and Engineering Faculty
      location: Brisbane, Australia
      email: stephanie.kobakian@qut.edu.au
      mark: 1
      author:
        - name: Stephanie Kobakian
    - name: Monash University
      department: Econometrics and Business Statistics Faculty
      location: Melbourne, Australia
      email: dicook@monash.edu
      mark: 1
      author:
        - name: Dianne Cook
keywords: ["statistics", "visual inference", "geospatial", "population"]
abstract: |
  The choropleth map display is commonly used for communicating spatial distributions across geographic areas. However, when choropleths are used the size of the geographic units will influence the understanding of the distribution derived by map users. The hexagon tile map is presented as an alternative display for visualizing population related distributions effectively. Visual inference is used to measure the power of the hexagon tile map design, and the choropleth is used as a comparison. The hexagon tile map display is tested using a distribution that is directly related to the geography, with values monotonically increasing from the North-West to South-East areas of Australia. This study finds in a hexagon tile map lineup the single map that contains a population related distribution is detected with greater probability than the same data displayed in a choropleth map. These findings should encourage map creators to implement alternative displays and consider a hexagon tile map when presenting spatial distributions of heterogeneous areas.
dev: pdf
bibliography: paper.bib
output:
  bookdown::pdf_book: 
      base_format: rticles::ieee_article
      number_sections: true
editor_options: 
  chunk_output_type: inline
link-citations: true
citation_package: biblatex
csl: springer-vancouver.csl
header-includes: \usepackage{graphicx}
---

<!-- Regular paper – 12 double column pages (Submissions may be up to 18 pages in length, subject to MOPC. All regular paper page limits include references and author biographies.) -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  error = FALSE, 
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  out.width = "100%")
```

@@

```{r data}
trend_colors <- c(
  "NW-SE" = "#B2DF8A",
  "Three Cities" = "#A6CEE3",
  "All Cities" = "#1F78B4")

@@

```{r reps}
replicate <- tibble(image_name = c("aus_cities_12_geo.png", "aus_cities_12_hex.png", 
                                   "aus_cities_3_geo.png", "aus_cities_3_hex.png",
                                   "aus_cities_4_geo.png", "aus_cities_4_hex.png",
                                   "aus_cities_9_geo.png", "aus_cities_9_hex.png",
                                   "aus_nwse_2_geo.png", "aus_nwse_2_hex.png",
                                   "aus_nwse_3_geo.png", "aus_nwse_3_hex.png",
                                   "aus_nwse_5_geo.png", "aus_nwse_5_hex.png",
                                   "aus_nwse_6_geo.png", "aus_nwse_6_hex.png",
                                   "aus_three_12_geo.png", "aus_three_12_hex.png",
                                   "aus_three_5_geo.png", "aus_three_5_hex.png",
                                   "aus_three_8_geo.png", "aus_three_8_hex.png",
                                   "aus_three_9_geo.png", "aus_three_9_hex.png"),
                    replicate = c(1, 1, 2, 2, 3, 3, 4, 4, 
                                  1, 1, 2, 2, 3, 3, 4, 4,
                                  1, 1, 2, 2, 3, 3, 4, 4))
# Add rep info to data
d <- d %>% left_join(., replicate, by = "image_name")
```

@@

Introduction
=============
<!-- General: comparison of displays, motivate hexagon tile map, enough info to motivate aims -->
<!-- geographies on choropleth -->
This study compares the effectiveness of the spatial display, a hexagon tile map, against the standard, a choropleth map, for communicating information about disease statistics. The choropleth map is the traditional method for visualizing aggregated statistics across administrative boundaries. The hexagon tile map builds on existing displays, such as the cartogram, and tessellated hexagon displays. A hexagon tile map forgoes the familiar boundaries, in favor of representing each geographic unit as an equally sized hexagon, placed approximately in the correct spatial location.  It differs in the relaxed requirement to have connected hexagons, and allows sparsely located hexagons. This type of display may be useful for other countries, and other purposes. The algorithm to construct a hexagon tile map is available in the R package sugarbag [@sugarbag].

@@

The Australian Cancer Atlas [@atlas] is an online interactive web tool created to explore the burden of cancer on Australian communities. There are many cancer types to be explored individually or aggregated. The Australian Cancer Atlas allows users to explore the patterns in the distributions of cancer statistics over the geographic space of Australia. It uses a choropleth map display and diverging color scheme to draw attention to relationships between neighboring areas. The hexagon tile map may be a useful alternative display to enhance the atlas.

@@

Background
=============

## Spatial data displays

@@

<!-- Many sets of Australian statistical areas have a large disparity between the smallest and largest units.
The rural communities in Australia operate on a much larger geographic space than small inner-city communities. When a spatial distribution is related to the size of the areas, or the population density, the size of the regions can allow erroneous conclusions to be drawn about the state of the statistic over the entire population, when presenting the statistics on a geographic map base. This occurs as large regions filled with a consistent colour or pattern can draw the attention of map readers, and small regions are not paid equal attention [@CTTMB]. An alternative display can be used to honestly communicate a spatial distribution for a set of heterogeneous areas [@NISCC]. 
This work aims to show that a hexagon tile map display is a viable alternative to the geographic map base for presenting population statistics.

A choropleth map is not the only display that can be used for presenting geospatial data. 
Alternative maps include various cartograms and tessellated tile maps. They allow other variables to be incorporated in the display to highlight the statistical values of various geographic areas.
-->

```{r liver, fig.height = 4, fig.width = 4, fig.cap = "A choropleth map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.", results = "asis"}
ggdraw() +
  draw_plot(rasterGrob(readPNG("figures/aus_liver_m.png")))
```

```{r liver-hex, fig.height = 4, fig.width = 4, fig.cap = "A hexagon tile map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.", results = "asis"}

ggdraw() +
  draw_plot(rasterGrob(readPNG("figures/aus_liver_m_hex.png")))
```
<!--
This will create the figure as needed in the paper, but I can't get the width argument to work properly.
\begin{figure*}[!t]
\centering
\includegraphics{paper_files/figure-latex/liver-1.pdf}
\caption{A choropleth map (a), and hexagon tile map (b) of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.}
\label{fig:liver2}
\end{figure*}
-->

@@

<!-- Hexagon tile map -->
The hexagon tile map algorithm, automatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag [@sugarbag]. Fig. \ref{fig:liver} shows the hexagon tile map, along with the choropleth map of liver cancer rates in Australia. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas have expanded, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is of interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

## Visual Inference

@@

```{r lineup, fig.cap = "This lineup of twelve hexagon tile map displays contains one map with a real population related structure. The rest are null plots that contain spatial correlation between neighbours.", fig.height = 8, fig.width = 8}
ggdraw() +
  draw_plot(rasterGrob(readPNG("figures/aus_cities_3_geo.png")))
```

@@

Methodology
============

@@

## Experimental factors

@@

The primary factor in the experiment is the plot type. The secondary factor is a trend model. Three trend models were developed, one mirroring a large spatial trend for which the choropleth would be expected to do well, and two with differing level of inner city hot spots. These latter two reflect the structure seen in the liver cancer data (Fig. \ref{fig:liver}). This produces six treatment levels:

@@

```{r exp-design, results = "asis", fig.width = 8, fig.height = 8, fig.cap = "The experimental design used in the visual inference study."}
ggdraw(xlim = c(0,1), ylim = c(0,1)) + draw_plot(rasterGrob(png::readPNG("figures/exp design chart.png")))
```

## Generating null data

@@

This smoothing allowed neighbors to be related to each other, but also allowed outliers, and showed distributions similar to the Liver cancer distribution (Fig. \ref{fig:liver}).

## Generating lineups

@@

One of the lineup locations was chosen to embed the real trend model map, in each of the four replicates, for the three trend models.
The location was chosen from a sub sample of the 12 possible locations. The chance of reptition using resampling was introduced to prevent participants from inducing the location by elimination, the locations 1, 7, 10 and 11 were not used.

@@

## Analysis

### Data Cleaning

@@

### Descriptive statistics

@@

Plots will be made in R [@RCore], with the `ggplot2` package [@ggplot2].

### Modelling

The likelihood of detecting the data plot in the lineup can be modelled using a linear mixed effects model. 
The R [@RCore] `glmer()` function in the `lme4` [@lme4] package implements generalised linear mixed effect models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

@@

where $y_{ij} = 0, 1$ is the logg odds for whether the subject detected the data plot, $\mu$ is the overall mean, $\tau_i, i=1,2$ is the map type effect, $\delta_j$ is the trend model effect. We are allowing for an interaction between map type and trend model as the response is binary, so a logistic model was used. As each participant provides results from 12 lineups, this model can account for each individual participants’ abilities as it includes a subject-specific random intercept.

@@

## Web application to collect responses

@@

## Participants

@@

## Demographic data collection

@@

<!--
## Experimental design

The choropleth map was used as the comparative visualization for presenting the lineups [@VVSIALM] as this is the common display for spatial cancer data.
Geographic distributions usually have some degree of spatial autocorrelation between neighbors.
This feature was incorporated in all maps shown in the lineup displays, the map that contained the trend feature shown in only one set of data was also affected by spatial autocorrelation.
A line up protocol was implemented to arrange $N$ maps in each display. 
A reasonable amount of null plots $N-1$ in the lineup was chosen to ensure the real data map was well hidden. A reasonable number of plots to show in each lineup, $N = 12$ was chosen to not overwhelm participants due to the detailed choropleth maps of Australian SA3 areas.

The hypotheses for each lineup are
$H_0$ : All plots look the same
$H_a$ : One plot looks different to the other plots

The same data was visualized on a choropleth map, and on a hexagon tile map, however in this study the participants only saw one of these two displays. The accuracy and times taken were aggregated for Groups and.
Comparing the results of participants who see the choropleth to those who see a hexagon tile map will show that population related distributions are spotted more frequently in a hexagon tile map display.

Let $n$ be the number of independent observers and $x_i$ the
number of observers who picked plot $i$, $i = \{1,...,m\}$

Then $x_i, x_2, ..., x_m$ follows a multinomial
distribution$Mult_{\pi_1, \pi_2, ...., \pi_m}(x_i, x_2, ..., x_m)$ with 
$\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an
observer, which we can estimate as $\hat{\pi}_i = x_i/n$.
The researchers compared the length of time taken, and the accuracy of the participants choices.
The power of a lineup can therefore be estimated as the ratio of correct
identifications $x$ out of $n$ viewings.

## The methods of data analysis used
The data analysis methods used in order to analyse and collate the results included downloading the survey submissions and opening them into the analysis software R [@RCore].

For each of the 12 lineup displays the researchers calculated:

 - accuracy: the proportion of subjects who detected the data plot
 - efficiency: average time taken to respond
-->

Results
============

<!--
The results section explores the [Accuracy] and contrasts the detection rate of the real data plot in the lineups.
The second section considers the [Speed] participants had when submitting their responses to the three questions asked regarding each display.
The [Certainty] participants felt about their decisions and the [Reason] for their choice in plot is also discussed.
-->

@@

## Participant demographics

@@

## Accuracy

Fig. \ref{fig:detect-compare} displays the average detection rates for the two types of plot separately for each trend model. Each trend model was tested using four repetitions, evaluations on the same data set were seen as either choropleths or hexagon tile maps by each group as specified in Table. \ref{fig:exp-design}; the detection rates for each display are connected by a line segment. The Three Cities and All Cities trend models shown in the hexagon tile map allowed viewers to detect the data plot substantially more often than the choropleth counterparts. One replicate for the All Cities group had similar detection rates for both plot types, the rate of detection using the choropleth map was much higher than other replicates. Surprisingly, participants could also detect the gradual spatial trend in the NW-SE group from the hexagon tile map. We expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs slightly better, or equally as well.

```{r detect-compare, fig.cap = "The detection rates achieved by participants are contrasted when viewing the four replicates of the three trend models. Each point shows the probability of detection for the lineup display, the facets separate the trend models hidden in the lineup. The points for the same data set shown in a choroleth or hexagon tile map display are linked to show the difference in the detection rate.", fig.height=5}
# Detectability rate for each lineup (image)
d_smry <- d %>% group_by(trend, type, replicate) %>%
  # pdetect measures the aggregated accuracy of the choices
  summarise(pdetect = length(detect[detect == 1])/length(detect)) %>%
  ungroup()

# Numerical summary
diffs <- d_smry %>% spread(type, pdetect) %>%
  mutate(dif = `Hex.` - `Choro.`)

# Plot summary
ggplot(d_smry, aes(x = type, y = pdetect, color = trend)) +
  geom_point(size = 2) +  
  geom_line(size = 1, aes(group = replicate)) +  
  facet_wrap(~trend) +
  scale_color_manual(values = trend_colors) +
  xlab("Type of areas visualized") +
  ylab("Detection rate") + 
  ylim(0,1) +
  guides(color = FALSE)
```

Table. \ref{tab:desc-stats} shows the means and standard deviations of the detection rate for each type of plot and each trend model. This also gives the standard deviations, the smallest standard deviation for all sets of replicates was the Three Cities trend model shown in a Choropleth display. This group of displays had a very small detection rate of 0.04. The mean detection rate for the Three Cities trend model shown as choropleth map lineups was also the smallest at 0.40.
The North-West to South-East (NW-SE) trend model unexpectedly had a higher mean detection rate for the hexagon tile map displays, but the difference in the means of detection rate was only 0.10.

@@

d %>% group_by(type, trend) %>%
  summarise(m = as.character(round(mean(detect), 2)),
      std.dev = as.character(round(sd(detect), 2))) %>% 
  mutate(std.dev = ifelse(std.dev == 0.5, "(0.50)", paste0("(", std.dev, ")"))) %>%   mutate(m = ifelse(m == 0.4, "0.40", m)) %>% 
  gather(stat, value, m, std.dev) %>% 
  pivot_wider(names_from = "trend", values_from = "value") %>% 
  arrange(type) %>% ungroup() %>% 
  mutate(Type = types) %>% select(Type, `NW-SE`, `Three Cities`, `All Cities`) %>%
  knitr::kable(., format = "latex", align = "lccc", booktabs = TRUE, 
    linesep = c("", "\\addlinespace"),
    caption = "The mean and standard deviation of the rate of detection for each trend model, calculated for the choropleth and hexagon tile map displays.")
```

@@

```{r detect-glmer1, results="asis"}
# Mixed effects models
glmer1 <- glmer(detect ~ type*trend + (1|contributor), 
              family = binomial, data = d)

@@

detection_rates <- broom.mixed::tidy(glmer1) %>%
  mutate(detection_rates = round(exp(estimate)/(1+exp(estimate)),2)) %>%
  select(term, estimate, detection_rates) %>% pull(detection_rates)

broom.mixed::tidy(glmer1) %>%
  mutate_at(.vars = c("estimate", "std.error"), round, 2) %>% 
  mutate(p.value = round(p.value, digits=2)) %>% 
  rowwise() %>%
  mutate(sig = case_when(
  p.value <= 0.001 ~ "$^{***}$",
  p.value <= 0.01 ~  "$^{**}$",
  p.value <= 0.05 ~  "$^{*}$",
  p.value <= 0.01 ~  "$^{.}$",
  TRUE ~ "$^{ }$")) %>%
  ungroup() %>% 
  filter(!is.na(std.error)) %>%
  mutate(term = glmer_terms) %>% 
  select(Term = term, 
    Est. = estimate, 
    Sig. = sig,
    `Std. Error` = std.error, 
    `P val` = p.value) %>% 
  knitr::kable(format = "latex", escape = FALSE, align= "rrlrr", 
    booktabs = T, linesep = c("", "\\addlinespace", "", "\\addlinespace", ""), 
    caption = "The model output for the generalised linear mixed effect model for detection rate. This model considers the type of display, the trend model hidden in the data plot, and accounts for contributor performance.")
```

## Speed

@@

```{r beeswarm, fig.cap = "The distribution of the time taken (seconds) to submit a response for each combination of trend, whether the data plot was detected, and type of display, shown using horizontally jittered dotplots. The colored point indicates average time taken for each plot type. Although some participants take just a few seconds per evaluation, and some take as much as mcuh as 60 seconds, but there is very little difference in time taken between plot types.", fig.height=8}
# Di playing
s <- d %>% group_by(type, trend, detect_f) %>%
  summarise(m=median(time_taken), 
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75))
library(ggbeeswarm)
ggplot() + 
  geom_quasirandom(data=d, aes(x=type, y=time_taken), alpha=0.9) + 
  #geom_hline(data=s, aes(yintercept=m, color=type)) +
  geom_point(data=s, aes(x=type, y=m, color=type), size=5, alpha=0.7) +
  #geom_errorbar(data=s, aes(x=type, ymin=q1, ymax=q3, color=type), width=0.3, size=2) +
  scale_color_brewer("", palette = "Dark2") +
  facet_grid(trend~detect_f) + 
  ylab("Time taken (seconds)") + xlab("") +
  theme(legend.position="bottom")
```

## Certainty

@@

```{r certainty, fig.cap = "The amount of times each level of certainty was chosen by participants when viewing hexagon tile map or choropleth displays. Participants were more likely to choose a high certainty when considering a Choropleth map. The mid value of 3 was the default certainty, it was chosen most for the Hexagon tile map displays.", fig.height=5}
d <- d %>% 
  mutate(certainty = as_factor(certainty)) %>% 
  mutate(replicate_f = as_factor(replicate))

@@

## Reason

@@

Acknowledgment
==============

The authors would like to thank the Australian Cancer Atlas team for discussions regarding alternative spatial visualizations, and Professor Kerrie Mengersen and Dr Earl Duncan for regular meetings filled with suggestions and comments. Mitchell O'Hara-Wild was a co-developer of the taipan [@taipan] R package for image tagging, used as the base for the web app constructed to collect participant evaluations of lineups. We are thankful for the NUMBATs (Non-Uniform Monash Business Analytics Team) for participating in the pilot study that helped to assess the experimental design and determine an appropriate sample size for the study.

The source code to produce this document can be found on [GitHub](https://github.com/srkobakian/experiment/tree/master/paper).
Supplementary materials have been included to discuss the survey procedures and the lineups that were used.
The full set of images can also be found at: https://github.com/srkobakian/experiment/tree/master/figures/final.

The supplementary material contains:

@@

The analysis of the work was completed in R [@RCore] with the use of the following packages:

>>>>

---
chapter: 4
knit: "bookdown::render_book"
---

# Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer {#experiment}

This chapter tests the performance of the hexagon tile map display created using the algorithm discussed in Section \ref{algorithm}.
It outlines the lineup protocol method of visual inference that can be used to test the effectiveness of information visualisations.
Using a two factor experimental design, the experiment contrasts the performance of participants when they viewed a choropleth map, and their performance when viewing a hexagon tile map.
The experiment also considered three types of spatial trends, one geographic trend, and two population related distributions. 
The results showed that participants did in fact more frequently find the population related distributions when using the hexagon tile map.

This chapter will be submitted to the journal *IEEE Transactions of Visualisation and Computer Graphics*.

## Abstract {-}

The choropleth map display is commonly used for communicating spatial distributions across geographic areas. However, when choropleths are used the size of the geographic units will influence the understanding of the distribution derived by map users. The hexagon tile map is presented as an alternative display for visualizing population related distributions effectively. Visual inference is used to measure the power of the hexagon tile map design, and the choropleth is used as a comparison. The hexagon tile map display is tested using a distribution that is directly related to the geography, with values monotonically increasing from the North-West to South-East areas of Australia. This study finds in a hexagon tile map lineup the single map that contains a population related distribution is detected with greater probability than the same data displayed in a choropleth map. These findings should encourage map creators to implement alternative displays and consider a hexagon tile map when presenting spatial distributions of heterogeneous areas.

```{r setup-4, echo=FALSE, message=FALSE, warning=FALSE, comment = FALSE}
library(knitcitations)
library(RefManageR)
library(sf)
library(sugarbag)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  error = FALSE, 
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  out.width = "80%")
options("citation_format" = "pandoc")
BibOptions(check.entries = FALSE, style = "markdown")
```

@@

```{r data-4}
trend_colors <- c(
  "NW-SE" = "#B2DF8A",
  "Three Cities" = "#A6CEE3",
  "All Cities" = "#1F78B4")

@@

```{r reps4}
replicate <- tibble(image_name = c("aus_cities_12_geo.png", "aus_cities_12_hex.png", 
                                   "aus_cities_3_geo.png", "aus_cities_3_hex.png",
                                   "aus_cities_4_geo.png", "aus_cities_4_hex.png",
                                   "aus_cities_9_geo.png", "aus_cities_9_hex.png",
                                   "aus_nwse_2_geo.png", "aus_nwse_2_hex.png",
                                   "aus_nwse_3_geo.png", "aus_nwse_3_hex.png",
                                   "aus_nwse_5_geo.png", "aus_nwse_5_hex.png",
                                   "aus_nwse_6_geo.png", "aus_nwse_6_hex.png",
                                   "aus_three_12_geo.png", "aus_three_12_hex.png",
                                   "aus_three_5_geo.png", "aus_three_5_hex.png",
                                   "aus_three_8_geo.png", "aus_three_8_hex.png",
                                   "aus_three_9_geo.png", "aus_three_9_hex.png"),
                    replicate = c(1, 1, 2, 2, 3, 3, 4, 4, 
                                  1, 1, 2, 2, 3, 3, 4, 4,
                                  1, 1, 2, 2, 3, 3, 4, 4))
# Add rep info to data
d <- d %>% left_join(., replicate, by = "image_name")
```

@@

## Introduction
<!-- General: comparison of displays, motivate hexagon tile map, enough info to motivate aims -->
<!-- geographies on choropleth -->
This study compares the effectiveness of the spatial display, a hexagon tile map, against the standard, a choropleth map, for communicating information about disease statistics. The choropleth map is the traditional method for visualizing aggregated statistics across administrative boundaries. The hexagon tile map builds on existing displays, such as the cartogram, and tessellated hexagon displays. A hexagon tile map forgoes the familiar boundaries, in favor of representing each geographic unit as an equally sized hexagon, placed approximately in the correct spatial location.  It differs in the relaxed requirement to have connected hexagons, and allows sparsely located hexagons. This type of display may be useful for other countries, and other purposes. The algorithm to construct a hexagon tile map is available in the R package sugarbag [@sugarbag].

@@

The Australian Cancer Atlas [@TACA] is an online interactive web tool created to explore the burden of cancer on Australian communities. There are many cancer types to be explored individually or aggregated. The Australian Cancer Atlas allows users to explore the patterns in the distributions of cancer statistics over the geographic space of Australia. It uses a choropleth map display and diverging color scheme to draw attention to relationships between neighboring areas. The hexagon tile map may be a useful alternative display to enhance the atlas.

@@

## Background

### Spatial data displays

@@

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{figures/aus_liver_m.pdf}
\caption{\label{fig:liver-geo}A choropleth map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{figures/aus_liver_m_hex.pdf}
\caption{\label{fig:liver-hex}A hexagon tile map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.}
\end{figure}

@@

<!-- Hexagon tile map -->
The hexagon tile map algorithm, automatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag [@sugarbag]. Fig. \ref{fig:liver-geo} shows the hexagon tile map, along with the choropleth map of liver cancer rates in Australia. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas have expanded, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is of interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

### Visual Inference

@@

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{figures/aus_cities_3_geo.png}
\caption{\label{fig:lineup}This lineup of twelve hexagon tile map displays contains one map with a real population related structure. The rest are null plots that contain spatial correlation between neighbours.}
\end{figure}

@@

### Methodology

@@

### Experimental factors

@@

The primary factor in the experiment is the plot type. The secondary factor is a trend model. Three trend models were developed, one mirroring a large spatial trend for which the choropleth would be expected to do well, and two with differing level of inner city hot spots. These latter two reflect the structure seen in the liver cancer data (Fig. \ref{fig:liver-geo}). This produces six treatment levels:

@@

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{figures/experiment_design.pdf}
\caption{\label{fig:exp-design}The experimental design used in the visual inference study.}
\end{figure}

### Generating null data

@@

This smoothing allowed neighbors to be related to each other, but also allowed outliers, and showed distributions similar to the Liver cancer distribution (Fig. \ref{fig:liver-geo}).

### Generating lineups

@@

One of the lineup locations was chosen to embed the real trend model map, in each of the four replicates, for the three trend models.
The location was chosen from a sub sample of the 12 possible locations. The chance of repetition using resampling was introduced to prevent participants from inducing the location by elimination, the locations 1, 7, 10 and 11 were not used.

@@

### Analysis

#### Data Cleaning

@@

#### Descriptive statistics

@@

Plots will be made in R [@R], with the `ggplot2` package [@ggplot2].

#### Modelling

The likelihood of detecting the data plot in the lineup can be modelled using a linear mixed effects model. 
The R [@R] `glmer()` function in the `lme4` [@lme4] package implements generalised linear mixed effect models. The model used includes the two main effects map type and trend model, which gives the fixed effects model to be:

@@

where $y_{ij} = 0, 1$ is the log odds for whether the subject detected the data plot, $\mu$ is the overall mean, $\tau_i, i=1,2$ is the map type effect, $\delta_j$ is the trend model effect. We are allowing for an interaction between map type and trend model as the response is binary, so a logistic model was used. As each participant provides results from 12 lineups, this model can account for each individual participants’ abilities as it includes a subject-specific random intercept.

@@

### Web application to collect responses

@@

### Participants

@@

### Demographic data collection

@@

## Results

@@

### Participant demographics

@@

### Accuracy

Fig. \ref{fig:detect-compare} displays the average detection rates for the two types of plot separately for each trend model. Each trend model was tested using four repetitions, evaluations on the same data set were seen as either choropleths or hexagon tile maps by each group as specified in Table. \ref{fig:exp-design}; the detection rates for each display are connected by a line segment. The Three Cities and All Cities trend models shown in the hexagon tile map allowed viewers to detect the data plot substantially more often than the choropleth counterparts.
One replicate for the All Cities group had a similar detection rate for both the choropleth and the hexagon tile map. Interestingly, in post-analysis we found that participants chose the data display in the choropleth lineup for reasons unrelated to the All Cities data structure.
Participants detected the gradual spatial trend in the NW-SE group equally well from both map types. This was a pleasant surprise; we expected that the choropleth map would be superior for the type of spatial pattern, but the data suggests the hexagon tile map performs equally as well.

```{r detect-compare, fig.cap = "The detection rates achieved by participants are contrasted when viewing the four replicates of the three trend models. Each point shows the probability of detection for the lineup display, the facets separate the trend models hidden in the lineup. The points for the same data set shown in a choroleth or hexagon tile map display are linked to show the difference in the detection rate.", fig.height=5}
## Detectability rate for each lineup (image)
d_smry <- d %>% group_by(trend, type, replicate) %>%
  ## pdetect measures the aggregated accuracy of the choices
  summarise(pdetect = length(detect[detect == 1])/length(detect)) %>%
  ungroup()

## Numerical summary
diffs <- d_smry %>% spread(type, pdetect) %>%
  mutate(dif = `Hex.` - `Choro.`)

## Plot summary
ggplot(d_smry, aes(x = type, y = pdetect, color = trend)) +
  geom_point(size = 2) +  
  geom_line(size = 1, aes(group = replicate)) +  
  facet_wrap(~trend) +
  scale_color_manual(values = trend_colors) +
  xlab("Type of areas visualized") +
  ylab("Detection rate") + 
  ylim(0,1) +
  guides(color = FALSE)
```

Table. \ref{tab:desc-stats} shows the means and standard deviations of the detection rate for each type of plot and each trend model. This also gives the standard deviations, the smallest standard deviation for all sets of replicates was the Three Cities trend model shown in a choropleth display. This group of displays had a very small detection rate of 0.04. The mean detection rate for the Three Cities trend model shown as choropleth map lineups was also the smallest at 0.40.
The North-West to South-East (NW-SE) trend model unexpectedly had a higher mean detection rate for the hexagon tile map displays, but the difference in the means of detection rate was only 0.10.

@@

d %>% group_by(type, trend) %>%
  summarise(m = as.character(round(mean(detect), 2)),
      std.dev = as.character(round(sd(detect), 2))) %>% 
  mutate(std.dev = ifelse(std.dev == 0.5, "(0.50)", paste0("(", std.dev, ")"))) %>%   mutate(m = ifelse(m == 0.4, "0.40", m)) %>% 
  gather(stat, value, m, std.dev) %>% 
  pivot_wider(names_from = "trend", values_from = "value") %>% 
  arrange(type) %>% ungroup() %>% 
  mutate(Type = types) %>% select(Type, `NW-SE`, `Three Cities`, `All Cities`) %>%
  knitr::kable(., format = "latex", align = "lccc", booktabs = TRUE, 
    linesep = c("", "\\addlinespace"),
    caption = "The mean and standard deviation of the rate of detection for each trend model, calculated for the choropleth and hexagon tile map displays.") %>% 
  kable_styling(latex_options =c("hold_position"))
```

@@

```{r detect-glmer1, results="asis"}
## Mixed effects models
glmer1 <- glmer(detect ~ type*trend + (1|contributor), 
              family = binomial, data = d)

@@

detection_rates <- tidy(glmer1) %>%
  mutate(detection_rates = round(exp(estimate)/(1+exp(estimate)),2)) %>%
  select(term, estimate, detection_rates) %>% pull(detection_rates)

tidy(glmer1) %>%
  mutate_at(.vars = c("estimate", "std.error"), round, 2) %>% 
  mutate(p.value = round(p.value, digits=2)) %>% 
  rowwise() %>%
  mutate(sig = case_when(
  p.value <= 0.001 ~ "$^{***}$",
  p.value <= 0.01 ~  "$^{**}$",
  p.value <= 0.05 ~  "$^{*}$",
  p.value <= 0.01 ~  "$^{.}$",
  TRUE ~ "$^{ }$")) %>%
  ungroup() %>% 
  filter(!is.na(std.error)) %>%
  mutate(term = glmer_terms) %>% 
  select(Term = term, 
    Est. = estimate, 
    Sig. = sig,
    `Std. Error` = std.error, 
    `P val` = p.value) %>% 
  knitr::kable(format = "latex", escape = FALSE, align= "rrlrr", 
    booktabs = T, linesep = c("", "\\addlinespace", "", "\\addlinespace", ""), 
    caption = "The model output for the generalised linear mixed effect model for detection rate. This model considers the type of display, the trend model hidden in the data plot, and accounts for contributor performance.") %>% 
  kable_styling(latex_options =c("hold_position"))
```

### Speed

@@

```{r beeswarm, fig.cap = "The distribution of the time taken (seconds) to submit a response for each combination of trend, whether the data plot was detected, and type of display, shown using horizontally jittered dotplots. The colored point indicates average time taken for each plot type. Although some participants take just a few seconds per evaluation, and some take as much as mcuh as 60 seconds, but there is very little difference in time taken between plot types.", fig.height=8}
## Di playing
s <- d %>% group_by(type, trend, detect_f) %>%
  summarise(m=median(time_taken), 
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75))
library(ggbeeswarm)
ggplot() + 
  geom_quasirandom(data=d, aes(x=type, y=time_taken), alpha=0.9) + 
  #geom_hline(data=s, aes(yintercept=m, color=type)) +
  geom_point(data=s, aes(x=type, y=m, color=type), size=5, alpha=0.7) +
  #geom_errorbar(data=s, aes(x=type, ymin=q1, ymax=q3, color=type), width=0.3, size=2) +
  scale_color_brewer("", palette = "Dark2") +
  facet_grid(trend~detect_f) + 
  ylab("Time taken (seconds)") + xlab("") +
  theme(legend.position="bottom")
```

### Certainty

@@

```{r certainty, fig.cap = "The amount of times each level of certainty was chosen by participants when viewing hexagon tile map or choropleth displays. Participants were more likely to choose a high certainty when considering a choropleth map, but more likely to be wrong for the All Cities and Three Cities patterns. Participants were less certain in their responses for the hexagon tile map lineups, perhaps reflecting the lack of familiarity.", fig.height=5}
d <- d %>% 
  mutate(certainty = as_factor(certainty)) %>% 
  mutate(replicate_f = as_factor(replicate))

@@

### Reason

@@

## Supplementary meaterial

The appendix \ref{appendix} contains:

@@
